Gradual Soundness: Lessons from Static Python

### 2021-12-01 pj-7.1/paper.tex
pj: intro cleanup, sec 2 thoughts

\documentclass[english,submission]{programming}

@@

\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{T}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptdyn}{\typefont{Dynamic}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{NoneType}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptfun}[2]{#1 \rightarrow #2}
\newcommand{\sptunion}[2]{#1 \cup #2}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptpydict}[2]{\paramtype{\typefont{PyDict}}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}

@@

\usepackage{cleveref}
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\usepackage[backend=biber]{biblatex}
\addbibresource{bg.bib}

@@

\authorrunning{K-C. Lu, B. Greenman, C. Meyer, D. Viehland, S. Krishnamurthi}
%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

@@

Unfortunately, the high run-time cost of sound types has split
the gradual typing community.
Industry teams have developed innovative type systems that accommodate
untyped designs, but these systems are all unsound [CITE].
These systems intentionally check nothing at run-time when untyped values enter
typed code.
Academic teams have primarily focused on the theory of sound
gradual types, formulating correctness properties and studying ever-more-descriptive types [CITE].
A few academics have studied the cost of run-time checks
in detail [CITE] and proposed implementation methods [CITE],
compiler technology [CITE],
and even weakened semantics [CITE], but none of these efforts
decisively close the performance gap.
The most promising attempt is the \emph{concrete} semantics
for gradual types [CITE kafka nom etc], but its low performance
overhead stems from severe restrictions on untyped code.
These restrictions may not be acceptable to the maintainers of a
working untyped codebase.
Indeed, the maintainers might prefer to re-implement their system
in a typed language, as Dropbox recently did with its core sync
engine.\footnote{\shorturl{https://}{dropbox.tech/infrastructure/rewriting-the-heart-of-our-sync-engine}}

@@

This paper reports on a remarkable exception to the rule among industry-made gradual type systems.
The \SP{} team at Instagram has developed a sound type system for a subset of
Python along with a runtime system designed to
capitalize on the benefits of sound types.
The language design is a variation of concrete types that lets programmers
begin with shallow claims about data structures and gradually opt in to deeper soundness.
Remarkably, the performance of Python is an upper bound on the observed performance of
Static Python.

@@

\begin{itemize}
  \item
    Evidence that gradual soundness delivers fast performance incrementally,
    as programmers migrate to sound types.
    \SP{} can be applied to a codebase with no refactoring and no performance overhead.
    After refactoring to sound types, \SP{} performs well on a microbenchmark
    suite and in production.
  \item
    A formal description of the \SP{} core language and an analysis of its soundness
    and optimizations.
    Soundness is not uniform across all types.
    Some are fully sound while others are constructor sound.
  \item
    In-depth discussions of the expressiveness limitations in \SP{} relative to
    idealized gradual type systems, and the programming discipline that Instagram
    follows in their gradually-sound codebase.
\end{itemize}

@@

We have written this paper with two audiences in mind.
First, we want to encourage system-builders to reproduce the
\SP{} language design in another context and measure performance.
Second, we want to lend focus to theoretical research on gradual typing.
The restrictions that \SP{} adopts
may be worthwhile assumptions for theory going forward.
On the flip side, future research may find a way to lift some of
the restrictions currently in \SP{}.

@@

Although this program is typed, a different untyped module is free
to import the function \code{f} and invoke it with any sort of
argument.
For this reason, the compiled version of \cref{fig:sp-example} must
include a few run-time casts:
\begin{itemize}
  \item
    The function \code{f} must check that its arguments match the \code{PyDict} domain annotation.
    Concretely, this check accepts any Python dictionary.
  \item
    Because the domain check does not validate the contents of an incoming dictionary,
    the dict access (\code{x["A"]}) must check that its result matches the function codomain.
\end{itemize}
These checks ensure a \emph{shape-level} notion of type soundness~(\cite{s:model}).
In short: if an expression has type \code{T} and reduces to a value, then the value
matches the top-level constructor, or shape, of the type.
The shape for a Python dictionary is merely \code{dict}.
Other types have deeper shapes; for example, \SP{} includes a \code{ChkDict}
type that predicts the keys and values in a special kind of dictionary.

Thanks to these checks, \SP{} can leverage type soundness to generate efficient code.
Because the call to \code{f} in \cref{fig:sp-example} appears in typed code,
there is no need to check that it sends type-correct inputs.
Thus, the compiled call uses a Cinder bytecode instruction to skip past the
domain check and immediately enter the function body.

@@

\subsubsection{Checked Types}

@@

@@

OO restrictions:

@@

\begin{figure}[t]
  \(
    \mftypeF{\sptype_0}
    \mfeq
    \left\{\begin{array}{ll}
      \sptpydict{}{} & \mbox{if $\sptype_0 = \sptpydict{\sptclass}{\sptclass}$}
    \\
      \sptype_0 & \mbox{otherwise}
    \end{array}\right.
  \)

@@

\begin{theorem}[Type Soundness]
  If\ \(\vdash \spexpr_0 : \sptype_0\)
  then one of the following holds:
  \begin{itemize}
    \item
      \(\spexpr_0 \sprred \spvalue_0
        \mbox{ and }
        \vdash \spvalue_0 : \mftypeF{\sptype_0}
      \)
    \item
      \(\spexpr_0\) diverges
    \item
      \(\spexpr_0 \sprred \sperror\)
  \end{itemize}
\end{theorem}

\begin{theorem}[Optimization Soundness]
  If\ \(\vdash \spexpr_0 : \mftypeF{\sptype_0}\)
  then\ \(\vdash \mfopt{\spexpr_0} : \mftypeF{\sptype_0}\)
\end{theorem}

\begin{theorem}[Optimization Effectiveness]
  FILL how do we know optimizations are useful rather than trivial?
  %% - length of eval(e) vs. eval(opt(e)) might be longer (good) or shorter (bad)
  %%   for general heuristic opts.
  %% - is [opt(e) != e] true for interesting programs?
\end{theorem}

@@

@@

Nom~\cite{mt-oopsla-2017} is foundational.

Full Monty(?)~\cite{pmmwplck-oopsla-2013} core calculus for Python language, foundational.

@@

\acks{
  Thanks to
   Guido van Rossom for stimulating tweets.
  This work was partly supported by the US National Science Foundation.
  This research was also developed with funding from the Defense Advanced Research Projects Agency (DARPA) and the Air Force Research Laboratory (AFRL).
  The views, opinions and/or findings expressed are those of the author and should not be interpreted as representing the official views or policies of the Department of Defense or the U.S.~Government.
  Greenman received support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
}

>>>>

\documentclass[english,cleveref,submission]{programming}

@@

\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{T}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptdyn}{\typefont{Dynamic}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{NoneType}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptfun}[2]{#1 \rightarrow #2}
\newcommand{\sptunion}[2]{#1 \cup #2}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptrawpydict}{\typefont{PyDict}}
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}

@@

%\usepackage{cleveref}
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\usepackage[backend=biber]{biblatex}
\addbibresource{bg.bib}

@@

%\authorrunning{K-C. Lu, B. Greenman, C. Meyer, D. Viehland, S. Krishnamurthi}

@@

Unfortunately, the high run-time cost of sound types has split
the gradual typing community.
Industry teams have developed innovative type systems that accommodate
untyped designs, but are unsound [CITE].
These systems intentionally check nothing at run-time when untyped values enter
typed code.
Academic teams have primarily focused on the theory of sound
gradual types, formulating correctness properties and studying ever-more-descriptive types [CITE].
A few academics have studied the cost of run-time checks
in detail [CITE] and proposed implementation methods [CITE],
compiler technology [CITE],
and even weakened semantics [CITE], but these efforts have not yet
decisively closed the performance gap.
The most promising attempt is the \emph{concrete} semantics
for gradual types [CITE kafka nom etc], but its low performance
overhead stems from severe restrictions on untyped code.
%% In other words, concrete types impose low performance costs
%% but high migration costs.
%% ... you can have any color as long as it's black -h.ford
%% ... concrete expresses few programs, but runs them all nicely
Whether developers can accept these restrictions is unclear;
indeed, they may prefer a whole-sale migration over an awkward
gradual transition.\footnote{Successful migrations of untyped codebases
to a typed language are rare, but not unheard of.
Twitter ported its server-side code from Ruby to Scala
and Dropbox moved its core sync engine from Python mypy to Rust: see
\shorturl{http://www.}{artima.com/scalazine/articles/twitter\_on\_scala.html} and
\shorturl{https://}{dropbox.tech/infrastructure/rewriting-the-heart-of-our-sync-engine}.}

@@

This paper reports on a remarkable exception to the rule among industry-made gradual type systems.
The \SP{} team at Instagram has developed a \emph{sound} type system for a subset of
Python along with a runtime system that uses soundness to drive optimizations.
The language design is a variation of concrete types that eases its restrictions.
Instead of asking programmers to migrate from untyped code to sound concrete types
in one leap, \SP{} offers an intermediate step via shallow types that enforce
only the top-level shape of values.
Programmers can begin with shallow claims and gradually dial up soundness.
On a suite of benchmarks and production modules, Static Python out-performs untyped
Python across the board.

@@

\begin{itemize}
  \item
    Evidence that gradual soundness delivers Python-level performance at a low migration cost
    and faster performance as programmers gradually incorporate sound types.
    \SP{} can be applied to a codebase with no refactoring and marginal performance overhead.
    After refactoring to sound types, \SP{} performs well on a microbenchmark
    suite and in production.
  \item
    A formal description of the \SP{} core language and an analysis of its soundness
    and optimizations.
    Soundness is not uniform across all types.
    Some are fully sound while others are constructor sound.
  \item
    Discussions of the expressiveness limitations in \SP{} relative to
    idealized gradual type systems, and the programming discipline that Instagram
    follows in their gradually-sound codebase.
\end{itemize}

@@

We have written this paper with two audiences in mind.
First, we want to encourage system-builders to reproduce the
\SP{} language design in another context and measure its performance.
Second, we want to lend focus to researchers.
Some of the restrictions that \SP{} adopts may be useful for
other languages.
Other restrictions might be lifted by future research.

@@

Although this program is typed, a different untyped module is free
to import the function \code{f} and invoke it with any sort of
argument.
For this reason, the compiled version of \cref{fig:sp-example} must
include a few run-time casts:
\begin{itemize}
  \item
    The function \code{f} must check that its arguments match the \code{PyDict} domain annotation.
    \SP{} inserts a check that accepts any Python dictionary.
  \item
    Because the domain check does not validate the elements of an incoming dictionary,
    the dict access (\code{x["A"]}) must check that its result matches the function codomain.
\end{itemize}
These checks ensure a \emph{shape-level} notion of type soundness~(\cref{s:model}).
In short: if an expression has type \code{T} and reduces to a value, then the value
matches the top-level constructor, or shape, of the type.
The shape for a Python dictionary is merely \code{dict}.
Other types have deeper shapes; for example, \SP{} includes a \code{ChkDict}
type that predicts the keys and values in a special kind of dictionary.

Thanks to these checks, \SP{} can leverage type soundness to generate efficient code.
Because the call to \code{f} in \cref{fig:sp-example} appears in typed code,
there is no need to check that it sends type-correct inputs.
Thus, the compiled call is optimized to use a Cinder bytecode instruction
that skips the domain check and immediately enters the function body.

@@

\subsubsection{Checked Types and Shallow Types}

@@

\subsubsection{Mixed-Typed Class Hierarchies}

Typed classes can inherit from untyped classes.
Untyped classes can inherit from typed classes.

Interesting, because StrongScript disallowed.
(But Nom allowed right?! Maybe allowed in a funny way, possibly to support their Dyn.)

\subsubsection{Types for Python}

Which types accommodate Python code? Which from PEP typing etc.

Surprisingly Unions are static only.
Except for unions to false because those are so common.

\subsubsection{Other}

More notes and surprises:

\begin{itemize}
  \item
    Optional(T) is the only union. Others turn into Dynamic.
  \item
    TBD
\end{itemize}

@@

@@

\begin{figure}[t]
  \(
    \mftypeF{\sptype_0}
    \mfeq
    \left\{\begin{array}{ll}
      \sptrawpydict & \mbox{if $\sptype_0 = \sptpydict{\sptclass}{\sptclass}$}
    \\
      \sptype_0 & \mbox{otherwise}
    \end{array}\right.
  \)

@@

\begin{theorem}[Type Soundness]
  If\ \(~\vdash \spexpr_0 : \sptype_0\)
  then one of the following holds:
  \begin{itemize}
    \item
      \(\spexpr_0 \sprred \spvalue_0
        \mbox{ and }
        \vdash \spvalue_0 : \mftypeF{\sptype_0}
      \)
    \item
      \(\spexpr_0\) diverges
    \item
      \(\spexpr_0 \sprred \sperror\)
  \end{itemize}
\end{theorem}

\begin{theorem}[Optimization Soundness]
  If\ \(~\vdash \spexpr_0 : \mftypeF{\sptype_0}\)
  then\ \(~\vdash \mfopt{\spexpr_0} : \mftypeF{\sptype_0}\)
\end{theorem}

%\begin{theorem}[Optimization Effectiveness]
%  FILL how do we know optimizations are useful rather than trivial?
%  %% - length of eval(e) vs. eval(opt(e)) might be longer (good) or shorter (bad)
%  %%   for general heuristic opts.
%  %% - is [opt(e) != e] true for interesting programs?
%\end{theorem}

@@

\begin{center}
  \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/LuKC1024}
\end{center}

@@

Nom~\cite{mt-oopsla-2017} and MonNom~\cite{mt-oopsla-2021} are foundational.

Full Monty~\cite{pmmwplck-oopsla-2013} core calculus for Python language, foundational for us.

@@

%\acks{
%  Thanks to
%  Guido van Rossum for stimulating tweets.
%  This work was partly supported by the US National Science Foundation.
%  This research was also developed with funding from the Defense Advanced Research Projects Agency (DARPA) and the Air Force Research Laboratory (AFRL).
%  The views, opinions and/or findings expressed are those of the author and should not be interpreted as representing the official views or policies of the Department of Defense or the U.S.~Government.
%  Greenman received support from NSF grant 2030859 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
%}
